---
title: "xgboost-code"
format: html
editor: visual
---
#Setup
```{r}
#| message: false
#| warning: false

library(tidymodels)
library(finetune)
library(vip)
library(xgboost)
library(ranger)
library(tidyverse)
library(doParallel)
```

#Load data
```{r}
corn <- read_csv("../data/training/weatheryearcombined.csv")
corn
```

```{r}
#corn %>% #looking at data/variables by year
  #filter(year == 2023)
```

```{r}
#corn %>% #looking at how many unique hybrids
  #distinct(hybrid)
```


#Pre-processing 
##Data split
```{r corn split}
set.seed(931735) #setting seed so its reproducible

corn_split <- initial_split(corn,
                            prop = .7, #just using 70/30 split
                            strata = yield_mg_ha) #stratifying by target variable
corn_split
```

```{r corn train}
corn_train <- training(corn_split)
corn_train
```

```{r corn test}
corn_test <- testing(corn_split)
corn_test
```

##Distribution of yield
```{r}
ggplot() +
  geom_density(data = corn_train, 
               aes(x = yield_mg_ha),
               color = "red") +
  geom_density(data = corn_test, 
               aes(x = yield_mg_ha),
               color = "blue") 
```

##Data processing with recipe
```{r corn recipe}
corn_recipe <- recipe(yield_mg_ha ~ ., data = corn_train) %>% #creating recipe for preprocessing
  step_rm(replicate, #removing these because I was getting an error since these are all non-numerical
          block,
          year,
          site,
          hybrid,
          month_planted,
          month_harvested)
corn_recipe
```

```{r corn prep}
corn_prep <- corn_recipe %>% #prepping recipe
  prep()
corn_prep
```

#Training
##Model specification
```{r}
xgb_spec <- 
  boost_tree(
   #number of boosting iterations
    trees = tune(), #optimizing hyperparameter
         #maxximum depth of each tree
    tree_depth = tune(),
             #minimum samples required to split a node
    min_n = tune(),
        #step size shrinkage for each boosting step
    learn_rate = tune()) %>%
        #specify engine 
    set_engine("xgboost") %>%
       #set mode
    set_mode("regression")

xgb_spec
```

##Cross validation
```{r}
set.seed(235)
resampling_foldcv <- vfold_cv(corn_train,
                              v = 10) #10 fold cross validation
resampling_foldcv
resampling_foldcv$splits[[1]]
```

##Hyperparameter grid
```{r}
xgb_grid <- grid_latin_hypercube(
  tree_depth(), #hyperparameters
  min_n(),
  learn_rate(),
  trees(),
  size = 100 #combinations of hyperparameters
)

xgb_grid
```

#Model tuning
```{r}
set.seed(76544)
registerDoParallel(cores = parallel::detectCores() - 1)

xgb_res <- tune_race_anova(object = xgb_spec,
                      preprocessor = corn_recipe,
                      resamples = resampling_foldcv,
                      grid = xgb_grid,
                      control = control_race(save_pred = TRUE))

stopImplicitCluster()

xgb_res$.metrics[[2]]
```

```{r}
plot_race(xgb_res)
```

#Selecting best model
##RMSE
```{r}
best_rmse <- xgb_res %>% #lowest rmse
  select_best(metric = "rmse")%>% 
  mutate(source = "best_rmse")

best_rmse
```
```{r}
best_rmse_pct_loss <- xgb_res %>% #best rmse within 1% loss
  select_by_pct_loss("min_n",
                     metric = "rmse",
                     limit = 1
                     )%>% 
  mutate(source = "best_rmse_pct_loss")

best_rmse_pct_loss
```
```{r}
best_rmse_one_std_err <- xgb_res %>% #best rmse within 1 SE
  select_by_one_std_err(metric = "rmse",
                        eval_time = 100,
                        trees
                        )%>% 
  mutate(source = "best_rmse_one_std_err")

best_rmse_one_std_err
```

##R2
```{r}
best_r2 <- xgb_res %>% #greatest R2
  select_best(metric = "rsq")%>% 
  mutate(source = "best_r2")

best_r2
```

```{r}
best_r2_pct_loss <- xgb_res %>% #lower R2 within 1% loss
  select_by_pct_loss("min_n",
                     metric = "rsq",
                     limit = 1
                     ) %>% 
  mutate(source = "best_r2_pct_loss")

best_r2_pct_loss
```

```{r}
best_r2_one_std_error <- xgb_res %>% #lowest R2 within 1 SE
  select_by_one_std_err(metric = "rsq",
                        eval_time = 100,
                        trees
                        ) %>%
  mutate(source = "best_r2_one_std_error")

best_r2_one_std_error
```

##Compare and finalize model
```{r}
best_rmse %>% 
  bind_rows(best_rmse_pct_loss, 
            best_rmse_one_std_err, 
            best_r2, 
            best_r2_pct_loss, 
            best_r2_one_std_error)
```
#Final Specification
```{r}
final_spec <- boost_tree(
  trees = best_rmse_pct_loss$trees,           # Number of boosting rounds (trees)
  tree_depth = best_rmse_pct_loss$tree_depth, # Maximum depth of each tree
  min_n = best_rmse_pct_loss$min_n,           # Minimum number of samples to split a node
  learn_rate = best_rmse_pct_loss$learn_rate  # Learning rate (step size shrinkage)
) %>%
  set_engine("xgboost") %>%
  set_mode("regression")

final_spec
```

#Validation
```{r}
set.seed(10)
final_fit <- last_fit(final_spec,
                corn_recipe,
                split = corn_split)

final_fit %>%
  collect_predictions()
```

#Evaluate on test set
```{r}
final_fit %>%
  collect_metrics()
```

#Evaluate on training set
```{r}
final_spec %>%
  fit(yield_mg_ha ~ .,
      data = bake(corn_prep, 
                  corn_train)) %>%
  augment(new_data = bake(corn_prep, 
                          corn_train)) %>% 
  rmse(yield_mg_ha, .pred) %>%
  bind_rows(
    
    
# R2
final_spec %>%
  fit(yield_mg_ha ~ .,
      data = bake(corn_prep, 
                  corn_train)) %>%
  augment(new_data = bake(corn_prep, 
                          corn_train)) %>% 
  rsq(yield_mg_ha, .pred))
```

#Predicted vs observed plot
```{r}
final_fit %>%
  collect_predictions() %>%
  ggplot(aes(x = yield_mg_ha,
             y = .pred)) +
  geom_point() +
  geom_abline() +
  geom_smooth(method = "lm") +
  scale_x_continuous(limits = c(0, 20)) +
  scale_y_continuous(limits = c(0, 20)) 
```

#Variable Importance
```{r}
final_spec %>%
  fit(yield_mg_ha ~ .,
         data = bake(corn_prep, corn_train)) %>% #There little change in variable improtance if you use full dataset
    vi() %>%
  mutate(
    Variable = fct_reorder(Variable, 
                           Importance)
  ) %>%
  ggplot(aes(x = Importance, 
             y = Variable)) +
  geom_col() +
  scale_x_continuous(expand = c(0, 0)) +
  labs(y = NULL)
```

